Large scale risks from pesticides in small streams

### 2016-10-02 article.tex
more revisions

\subsection{Effect of precipitation on pesticide exposure}
The spatio-temporal intersection revealed that 5\% of the samples were taken at or after days with rainfall events greater than 10mm / day (Supplemental Figure~S6).
$prec_{0}$ and $prec_{-1}$ increased the probability of exceeding LOQ and the mean value of RQ.
$prec_{-1}$ had a stronger effect than $prec_{0}$. 
This difference was less pronounced for the mean value RQ (Figure~\ref{fig:fig5}, top).

The first quarter showed the lowest RQ and probability of exceeding LOQ.
Both increased during summer months and decreased towards the end of the year.
The differences were less pronounced for the value RQ and with higher variation (Figure~\ref{fig:fig5}, bottom).

@@

\subsection{Pesticides in small water bodies}
The dataset comprised 12,710 samples from 1,295 small water bodies.
In 346 (26.7\%) of these streams RQ were exceeded.
In total 1,173 samples (0.3\% of all and 5.6\% of samples with detects) showed RQ higher than 1 were observed.
Neonicotinoid insecticides and Chlorpyrifos showed the highest RQ exceedances (Figure~\ref{fig:fig6}).
For Thiacloprid, Imidacloprid and Chlorpyrifos the RAC was less than LOQ, therefore, all detections have a RQ~\textgreater~1. 
The herbicides Nicosulfuron and Diflufenican, as well as the fungicide Dimoxystrobin also showed high exceedances of RQ (29.5, 14.2 and 21.4 \% of all samples with detects).
The highest RQ were observed for Chlorpyrifos (max(RQ) = 244), Dimoxystrobin(max(RQ) = 117) and Isoproturon (max(RQ) = 80). 
Metabolites were most commonly detected if analysed (for example, Metazachlor sulfonic acid was detected in 82\% of all samples were it was analysed, see also Supplemental Figure~S9).
Glyphosate was the compound with the highest detection rates, followed by Boscalid and Isoproturon. 
However, only the latter showed RQ exceedances (Figure~\ref{fig:fig6}).
In 44.8\% of samples more than one compound was quantified, with a maximum of 54 different compounds in one sample (Supplemental Figure~S8).

@@

%% -------------------------------------------------------------------------
\section{Discussion}
\subsection{Overview on the compiled dataset}
The compiled dataset of governmental monitoring data represents currently the most comprehensive one available for Germany.
Similar nationwide datasets have been compiled for the Netherlands \citep{vijver_spatial_2008}, Switzerland \citep{munz_pestizidmessungen_2011} and the United States (Water Quality Portal (WQP) \url{www.waterqualitydata.us}).
The data compiled and analysed here for Germany is of similar quantity and quality.

% current problems in monitoring and possible solutions
Nevertheless, a nationwide assessment of pesticide pollution is hampered by the inhomogeneity of monitoring data between federal states:
There are not only big differences in the spatial distribution and quantity of sampling sites (Figure~\ref{fig:fig1}), but also the spectrum of analyzed compounds (Figure~\ref{fig:fig2}) and differences in the quality of chemical analyses. 
For Thiacloprid, Imidaclorpid and Chlorpyrifos LOQ were above RAC.
For these compounds a lowering of LOQ is essential for reliable assessment.
Moreover, would a nation-wide assessment benefit from a harmonized spectrum of analysed compounds between federal states.

Given their high abundance in the landscape \citep{nadeau_hydrological_2007} SWB are underrepresented in the current monitoring (Figure~\ref{fig:fig3}). 
Especially, streams below 10~km\textsuperscript{2} are missing, which could be attributed to the missing categorisation in the WFD.

\subsection{Influence of agricultural land use and catchment size}
% agriculture
We found a strong influence of agriculture on the pollution of streams.
If there is more the 25\% agriculture within a catchment pesticides, it is likely that RAC will be exceeded with an increase in fully agricultural catchments (above 75 \% agriculture).
To our knowledge this is the first study investigating such thresholds of pesticide exposure.
Thresholds for agricultural land use have only been investigated in the past for biological communities.
\citet{feld_response_2013} found change points of biological community metric for agricultural land use at 40\% in low lands. 
Similarly, \citet{waite_agricultural_2014} found a threshold for aquatic diatoms at 40\%.
Our results coincide with these thresholds and suggest that pesticides might contribute to the observed biological changes.

% size
We did not find a relationship between pesticide pollution and catchment size.
However, previous studies have shown that SWB are more polluted that bigger streams \citep{schulz_field_2004,stehle_pesticide_2015,knauer_pesticides_2016}.
This could be explained by the relatively short gradient of catchment sizes in our dataset, with most of the streams being \textless $100~km^2$ (Figure~\ref{fig:fig3}, top).
For example the gradient of \citet{schulz_field_2004} covered 6 orders of magnitude.
Another reason might be the unequal distribution of catchment sizes, with fewer sites below $10~km^2$ and above $100~km^2$ catchment size (Figure~\ref{fig:fig3}, top).

\subsection{Effect of precipitation on pesticide exposure}
Our results revealed that pesticide sampling for chemical monitoring in Germany is mainly performed when no precipitation occurs. 
Nevertheless, we found higher RQ if samples where taken after rainfall events. 
Samples taken at the day of a rainfall-event showed less risk, than samples taken one day after a rainfall-event.
This could be explained by a sampling just before the actual rainfall event.
Pesticide concentrations in agricultural SWB generally show short term peak concentrations \citep{wittmer_significance_2010}, therefore, a sampling at the day after a rainfall-event might miss peak concentrations.
The effects of precipitation were more pronounced for the probability to exceed LOQ, with smaller effect sizes for RQ.
This could be attributed to a higher variability of absolute concentrations.
Overall, our results indicate that current pesticide monitoring strongly underestimates pesticide exposure.
Automatic event-drive samplers \citep{stehle_probabilistic_2013} and passive samplers \citep{fernandez_calibration_2014,moschet_evaluation_2015} may help overcome these shortcomings and provide a better representation, especially for SWB \citep{lorenz_specifics_2016}.

We found highest the probability of exceeding LOQ during summer (9.9\% for Q2) and lowest in the first Quarter (2.8\%, Figure~\ref{fig:fig5}, bottom right).
This yearly pattern coincides with their main application season for pesticides.
Nevertheless, there are compound specific differences in the yearly pattern, which explain the wide CI for the absolute RQ (Figure~\ref{fig:fig5}, bottom left).
For example, the herbicide Diflufenican showed highest RQ during the winter quarters (Supplemental Table~S4), which is the application period it is registered for in Germany \citep{bvl_online_2016}.
Our study suggest that pesticide exposure shows compound specific spatio-temporal dynamics.
Currently, little is known on these and further research on those might provide useful information for future ecological risk assessment.

\subsection{Pesticides in small water bodies}
Our results suggest that SWB are frequently exposed to biologically relevant pesticide concentrations.
\citet{stehle_pesticide_2015} found the highest percentage of RAC exceedances for organophosphate insecticides. 
By contrast, our results revealed that neonicotinoid insecticides show high exceedances, followed by the organophosphate chlorpyrifos. 
This difference can be attributed to the low sample size for neonicotinoid insecticides in their study (n = 33) compared to the dataset presented here (between 1,290 and 2,113 samples in SWB, Figure~\ref{fig:fig6}). 
Our results shows that this particular class of insecticides may currently pose a high risk to freshwater ecosystems.
Compared to \citet{stehle_pesticide_2015} we found much lower rates RAC exceedances (0.3\% of 372,304 samples with RAC vs 44\% of 1,566 samples). 
This can be attributed to different aims of the data sources: scientific research aims at finding pollutants, whereas monitoring aims mainly at surveillance of water quality, also during periods with lower pesticide usage and at natural sites. 
Contrary, \citet{knauer_pesticides_2016} found exceedances from monitoring data mainly for herbicides and fungicides and only one insecticide Chlorpyrifos.
This might reflect differences in pesticide use between countries and defined RACs.

From the definition of RAC it follows that if the concentration of a compound exceeds its RAC ecological effects are expected.
Accordingly, \citet{stehle_agricultural_2015} found that biological diversity is significantly reduced at a RQ of 1.12.
We found RQ values greater than 1.12 at 25\% of all SWB sites. 
Consequently, we conclude that pesticides from agricultural land use are a major treat to small water bodies, the biodiversity they host and the services they provide. 
Additional treat arises when taking into account that most pesticides do not occur individually but in mixtures \cite{schreiner_pesticide_2016} and may co-occur with other stressors \citep{schafer_contribution_2016}.

% Approval / Risk Assessment
Monitoring data, despite the outlined limitations, provides an opportunity to study large scale environmental occurrences of pesticides.
Nevertheless, such nationwide compilations, may not only be used for governmental surveillance, but also to answer other questions, like validation of exposure modeling \cite{knabel_fungicide_2014}, retrospective evaluation of regulatory risk assessment \citep{knauer_pesticides_2016,stehle_pesticide_2015}or occurrences of pesticide mixtures \cite{schreiner_pesticide_2016}. 
The high exceedances of RAC indicate that the approval process for pesticides must be checked and refined.

>>>>

\subsection{Effect of precipitation on pesticide exposure}
The spatio-temporal intersection revealed that 5\% of the samples were taken at or after days with rainfall events greater than 10mm / day (Supplemental Figure~S6).
$prec_{0}$ and $prec_{-1}$ increased the probability of exceeding LOQ and the mean value of RQ.
Precipitation before sampling ($prec_{-1}$) had a stronger effect than precipitation during sampling ($prec_{0}$). 
This difference was less pronounced for the mean value of RQ (Figure~\ref{fig:fig5}, top).

The first quarter showed the lowest RQ and probability of exceeding LOQ.
Both increased during summer months and decreased towards the end of the year.
The differences were less pronounced for the value of RQ and with higher variation (Figure~\ref{fig:fig5}, bottom).

@@

\subsection{Pesticides in small water bodies}
The dataset comprised 12,710 samples from 1,295 small water bodies.
In 346 (26.7\%) of these streams RQ were exceeded.
In total 1,173 samples (0.3\% of all and 5.6\% of samples with detects) showed RQ higher than 1 were observed.
Neonicotinoid insecticides and Chlorpyrifos showed the highest RQ exceedances (Figure~\ref{fig:fig6}).
For Thiacloprid, Imidacloprid and Chlorpyrifos the RAC was less than LOQ, therefore, all detections have a RQ~\textgreater~1. 
The herbicides Nicosulfuron and Diflufenican, as well as the fungicide Dimoxystrobin also showed high exceedances of RQ (29.5, 14.2 and 21.4 \% of all samples with detects).
The highest RQ were observed for Chlorpyrifos (max(RQ) = 244), Dimoxystrobin(max(RQ) = 117) and Isoproturon (max(RQ) = 80). 
Where analysed, metabolites exhibited the highest detection rates (for example, Metazachlor sulfonic acid was detected in 82\% of all samples were it was analysed, see also Supplemental Figure~S9).
Glyphosate was the compound with the highest detection rates, followed by Boscalid and Isoproturon. 
However, only the latter showed RQ exceedances (Figure~\ref{fig:fig6}).
In 44.8\% of samples more than one compound was quantified, with a maximum of 54 different compounds in one sample (Supplemental Figure~S8).

@@

%% -------------------------------------------------------------------------
\section{Discussion}
\subsection{Overview on the compiled dataset}
The compiled dataset of governmental monitoring data, with a particular focus on small streams, represents currently the most comprehensive one available for Germany.
Similar nationwide datasets have been compiled for the Netherlands \citep{vijver_spatial_2008}, Switzerland \citep{munz_pestizidmessungen_2011} and the United States (Water Quality Portal (WQP) \url{www.waterqualitydata.us}).
The data compiled and analysed here for Germany is of similar quantity and quality.

% current problems in monitoring and possible solutions
Nevertheless, a nationwide assessment of pesticide pollution is hampered by the inhomogeneity of monitoring data between federal states:
Beside large differences in the spatial distribution and quantity of sampling sites (Figure~\ref{fig:fig1}), the spectrum of analyzed compounds (Figure~\ref{fig:fig2}) and the quality of chemical analyses differed between states. 
For Thiacloprid, Imidaclorpid and Chlorpyrifos the LOQs were above the RAC, which means that some exceedances were likely not detected.
For these compounds a lowering of LOQ is essential for reliable assessment.
Moreover, a nation-wide assessment would benefit from a harmonized spectrum of analysed compounds between federal states.

Given their high abundance in the landscape \citep{nadeau_hydrological_2007} SWB are underrepresented in the current monitoring (Figure~\ref{fig:fig3}). 
Especially, sampling sites in streams below 10~km\textsuperscript{2} are disproportionally low, which may be attributed to the missing categorisation in the WFD.

\subsection{Influence of agricultural land use and catchment size}
% agriculture
We found a strong influence of agriculture on the pollution of streams.
If there is more the 25\% agriculture within a catchment pesticides, it is likely that a RAC will be exceeded, with a further increase the likelihood of RAC exceedance in fully agricultural catchments (above 75 \% agriculture).
To our knowledge this is the first study investigating such thresholds of pesticide risk.
Previous studies examined thresholds for \%agricultural land use with respect to the response of biological communities, integrating different agricultural stressors.
\citet{feld_response_2013} found change points of biological community metrics at 40\% agricultural land use  in low land streams.
Similarly, \citet{waite_agricultural_2014} found a threshold for aquatic diatoms at 40\%.
Our results coincide with these thresholds and suggest that pesticides might contribute to the observed biological changes.

% size
We did not find a relationship between pesticide pollution and catchment size.
However, previous studies showed that SWB are more polluted that bigger streams \citep{schulz_field_2004,stehle_pesticide_2015,knauer_pesticides_2016}.
This can be explained by the relatively short gradient of catchment sizes in our dataset, with most of the streams with catchments above $10~km^2$ and below $100~km^2$ (Figure~\ref{fig:fig3}, top).
For example the gradient of \citet{schulz_field_2004} covered 6 orders of magnitude.

\subsection{Effect of precipitation on pesticide exposure}
Our results revealed that pesticide sampling for chemical monitoring in Germany is mainly performed when no precipitation occurs. 
Nevertheless, we found higher RQ if samples where taken after rainfall events. 
Samples taken at the day of a rainfall-event showed less risk, than samples taken one day after a rainfall-event.
This could be explained by a sampling just before the actual rainfall event.
Pesticide concentrations in agricultural SWB generally show short term peak concentrations \citep{wittmer_significance_2010}, therefore, a sampling at the day after a rainfall-event might miss peak concentrations.
The effects of precipitation were more pronounced for the probability to exceed LOQ, with smaller effect sizes for RQ.
This could be attributed to a higher variability of absolute concentrations.
Overall, our results indicate that current pesticide monitoring relying on grab sampling, largely disconnected from precipitation events, underestimates pesticide risks.
Automatic event-drive samplers \citep{stehle_probabilistic_2013} and passive samplers \citep{fernandez_calibration_2014,moschet_evaluation_2015} may help overcome these shortcomings and provide a better representation, especially for SWB \citep{lorenz_specifics_2016}.

We found highest the probability of exceeding LOQ during summer (9.9\% for Q2) and lowest in the first Quarter (2.8\%, Figure~\ref{fig:fig5}, bottom right).
This yearly pattern coincides with their main application season for pesticides in Central Europe.
Nevertheless, there are compound specific differences in the yearly pattern, which explain the wide CI for the absolute RQ (Figure~\ref{fig:fig5}, bottom left).
For example, the herbicide Diflufenican showed highest RQ during the winter quarters (Supplemental Table~S4), which is the application period it is registered for in Germany \citep{bvl_online_2016}.
Our study suggests that pesticide risks display compound specific spatio-temporal dynamics.
Currently, little is known on these and further research on those might provide useful information for future ecological risk assessment.

\subsection{Pesticides in small water bodies}
Our results suggest that SWB are frequently exposed to ecologically relevant pesticide concentrations.
\citet{stehle_pesticide_2015} found the highest percentage of RAC exceedances for organophosphate insecticides. 
By contrast, our results that neonicotinoid insecticides showed highest exceedances of RACs, followed by the organophosphate chlorpyrifos. 
This difference can be attributed to the low sample size for neonicotinoid insecticides in their study (n = 33) compared to the dataset presented here (for example 2,113 samples of Thiacloprid, Figure~\ref{fig:fig6}). 
Overall, our results suggest that neonicotinoids may currently pose a high risk to freshwater ecosystems. 
Compared to \citet{stehle_pesticide_2015} we found much lower rates RAC exceedances (0.3\% of 372,304 samples with RAC vs 44\% of 1,566 samples). 
This can be attributed to different aims of the data sources: scientific research aims at finding pollutants, whereas monitoring aims mainly at surveillance of water quality, also during periods with lower pesticide usage and at natural sites. 
By contrast, \citet{knauer_pesticides_2016} found exceedances from monitoring data mainly for herbicides and fungicides and only one insecticide Chlorpyrifos.
This might reflect differences in pesticide use between countries and defined RACs.

From the definition of RAC it follows that if the concentration of a compound exceeds its RAC ecological effects are expected.
Indeed, \citet{stehle_agricultural_2015} found that biological diversity of stream invertebrates was significantly reduced at a RQ of 1.12.
We found RQ values greater than 1.12 at 25\% of all SWB sites. 
Consequently, we conclude that agricultural pesticides are a major threat to small streams, the biodiversity they host and the services they provide. 
This threat may exacerbate because esticides often occur in mixtures \cite{schreiner_pesticide_2016} and may co-occur with other stressors \citep{schafer_contribution_2016}.

% Approval / Risk Assessment
Monitoring data, despite the outlined limitations, provides an opportunity to study large scale environmental occurrence patterns of pesticides.
Nevertheless, such nationwide compilations, may not only be used for governmental surveillance, but also to answer other questions, like validation of exposure modeling \cite{knabel_fungicide_2014}, retrospective evaluation of regulatory risk assessment \citep{knauer_pesticides_2016,stehle_pesticide_2015}or occurrences of pesticide mixtures \cite{schreiner_pesticide_2016}, though the sampling design need to account for precipitation events to provide robust data. 
The high exceedances of RAC indicate that the authorisation process for pesticides may require further development.
